Lossy compression of DNA sequencing quality data

### 2015-04-20 manuscript/q-compression.tex
Added related work into methods section.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each class. In addition to
quantifying compression efficiency, we also assess the effects of
quality value information loss resulting from compression on
subsequent genomic analyses including read preprocessing (filtering
and trimming), genome assembly, and read mapping.

@@

@@

@@

\textit{Stapylococcus aureus} USA300, which are generated from a
fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
The corresponding reference sequence was obtained from the NCBI RefSeq
database (NC\_010063.1, NC\_010079.1, NC\_012417.1).

@@

We evaluate effects of information loss from quality value compression
on quality control steps of read filtering and trimming, which were
performed using Sickle~\citep{sickle}, quantify read characteristics
with PrinSeq~\citep{Schmieder:2011gd}, and make comparison to
uncompressed data.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the number
  of reads that are kept or discarded after applying compression with
  respect to the original, uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
  \cline{2-12}
& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
\cline{2-12}
 & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
\end{tabular}

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
\cline{2-12}
& proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
\end{tabular}

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
 & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
\cline{2-12}
& proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
\end{tabular}
\end{table*}

>>>>

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each class. ~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set.  We build on these prior works and access the effects of
% quantifying compression efficiency, we also assess the effects of
quality value information loss resulting from compression on additional
subsequent genomic analyses including read preprocessing (filtering
and trimming), genome assembly, and read mapping.

@@

\cite{Wan:2012kq} provide three similar lossy compression strategies based on binning the base error probability distribution: UniBinning, Truncating, and LogBinning.
UniBinning evenly splits the error probability distribution into a user-defined number of partitions.
Truncating treats a user-defined number of highest quality values as a single bin.
LogBinning works similar to UniBinning, except with the \emph{log} of the error probability distribution, which effectively bins the \textsc{ascii} quality values evenly.
Our 2-bin encoding is a combination of LogBinning and Truncating in that we are placing the highest quality values (as defined above) of the log of the error probability distribution into a single bin.

@@

\textsc{QualComp} is a lossy compression tool that models the quality sequence as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
Once the model parameters are calculated, they are stored by the decoder to later reconstruct a \emph{representative} quality value.
\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position in a quality sequence while minimizing the \textsc{mse}.
The quality sequences can be clustered before hand to produce more accurate models.

% they are stored for use by the decoder, and used to derive the number of bits used to encode each quality value, by minimizing the mean square error of a rate distortion problem. This optimization problem is parameterized by a user input, which is the total number of output bits the quality scores can consume, expressed as a rate r. The solution to the problem specifies, for position i in a read, the number of bits, 􏰗i, that should be used to encode the quality value in position i. For each quality value, if it is to be coded in 􏰗i bits, the Gaussian for that position is split into 2􏰗i regions of equal probability, and a bit pattern selected to indicate the region in which the quality value lies. The decoder reads in the Gaussian parameters and r, performs the optimization to recover the 􏰗i values and then emits a representative correspond- ing to the region specified by each code word.

@@

% Will be moved to supplemental.
%\textit{Stapylococcus aureus} USA300, which are generated from a
%fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
%and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
%The corresponding reference sequence was obtained from the NCBI RefSeq
%database (NC\_010063.1, NC\_010079.1, NC\_012417.1).

@@

We evaluate effects of information loss from quality value compression
on quality control steps of read filtering and trimming, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include N50, which is defined as the median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$ exceeds half of the assembly size) and corrected N50, which is the recalculated N50 size after the contigs are broken apart at locations of errors.  The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

% \begin{table*}[!tbhp]
% \centering
% \caption[]{Quality filtering/trimming for the uncompressed and
%   decompressed \textsc{fastq} files. Numbers correspond to the number
%   of reads that are kept or discarded after applying compression with
%   respect to the original, uncompressed values.}
% \label{tab:quality_control}
% \begin{tabular}{lr|cc|cc|cc|cc|cc}
%  & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
%  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
%   \cline{2-12}
% & kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
% {\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
% \cline{2-12}
%  & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
% \end{tabular}

% \bigskip

% \begin{tabular}{lr|cc|cc|cc|cc|cc}
%  & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
%  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
% \cline{2-12}
% & kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
% {\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
% \cline{2-12}
% & proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
% \end{tabular}

% \bigskip

% \begin{tabular}{lr|cc|cc|cc|cc|cc}
% &  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
%  & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
% \cline{2-12}
% & kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
% {\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
% \cline{2-12}
% & proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
% \end{tabular}
% \end{table*}
